# AI Impact on Jobs 2030 - Statistical Analysis

## Notebook 2 : Statistical Analysis

Following the descriptive exploration of the AI Impact on Jobs 2030 dataset, this notebook delves deeper into **statistical analyses** to better understand patterns related to automation risk.

We focus on key job characteristics such as salary, experience, education, and AI exposure, and explore their relationships with automation probability. This includes examining distributions, testing statistical assumptions, and applying appropriate analyses to summarize and compare groups.

The main guiding question remains:

**"Which types of jobs are most at risk of automation, and how do job characteristics are distributed across different levels of automation risk?"**

By rigorously examining these variables, we aim to provide clearer insights into the factors associated with automation vulnerability across occupations.

Author: J-F Jutras\
Date: December 2025\
Dataset: AI Impact on Jobs 2030 — Kaggle


```{r Data_Loading_and_Overview, echo = FALSE, results = 'asis', warning = FALSE, message = FALSE}

library(readr)

df <- read_csv("AI_Impact_on_Jobs_2030.csv", show_col_types = FALSE)

library(kableExtra)

table_style <- function(kbl) {
  kbl %>%
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left",
      font_size = 14
    ) %>%
    row_spec(
      0,
      bold = TRUE,
      background = "#1F4E79",
      color = "white"
    ) %>%
    column_spec(1, bold = TRUE)
}


```


## Automation Probability Across Education Levels


```{r anovaprobedu, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)

# -------------------------
# ANOVA Automation Probability ~ Education Level
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Education_Level, data = df)

# Résumé tidy de l'ANOVA
anova_summary <- broom::tidy(anova_model) %>%
  mutate(
    Sum_Sq = round(sumsq, 3),
    Mean_Sq = round(meansq, 3),
    F_value = round(statistic, 3),
    P_value = signif(p.value, 3)
  ) %>%
  select(term, Sum_Sq, Mean_Sq, F_value, P_value)

# Affichage avec ton style
kable(
  anova_summary,
  caption = "<b>ANOVA: Automation Probability by Education Level</b>",
  escape = FALSE
) %>%
  table_style()

```


```{r normaliteeduauto, echo = FALSE, message = FALSE, warning = FALSE }
library(ggplot2)
library(gridExtra)

# -------------------------
# Modèle ANOVA
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Education_Level, data = df)
residuals_anova <- residuals(anova_model)

# -------------------------
# 1. QQ-plot des résidus pour la normalité
# -------------------------
qq_plot <- ggplot(data.frame(residuals = residuals_anova), aes(sample = residuals)) +
  stat_qq(color = "#1F4E79") +
  stat_qq_line(color = "#A61C3C", linewidth = 1) +
  labs(title = "QQ Plot of Residuals") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# -------------------------
# 2. Plot des résidus vs valeurs ajustées pour l'hétéroscédasticité
# -------------------------
fitted_values <- fitted(anova_model)
resid_plot <- ggplot(data.frame(Fitted = fitted_values, Residuals = residuals_anova), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "#1F4E79", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# -------------------------
# Affichage côte à côte
# -------------------------
grid.arrange(qq_plot, resid_plot, ncol = 2)

```


```{r shapiroeduauto, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(car)       # pour LeveneTest
library(knitr)
library(kableExtra)

# -------------------------
# Modèle ANOVA
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Education_Level, data = df)
residuals_anova <- residuals(anova_model)

# -------------------------
# 1. Test de normalité des résidus (Shapiro-Wilk)
# -------------------------
shapiro_res <- shapiro.test(residuals_anova)
shapiro_df <- data.frame(
  Test = "Shapiro-Wilk (Residuals)",
  Statistic = round(shapiro_res$statistic, 3),
  P_value = signif(shapiro_res$p.value, 3)
)

# -------------------------
# 2. Test d'homogénéité des variances (Levene)
# -------------------------
# LeveneTest de car : centré sur la médiane
levene_res <- car::leveneTest(Automation_Probability_2030 ~ Education_Level, data = df, center = median)
levene_df <- data.frame(
  Test = "Levene (by Education Level)",
  Statistic = round(levene_res$`F value`[1], 3),
  P_value = signif(levene_res$`Pr(>F)`[1], 3)
)

# -------------------------
# 3. Combiner les deux résultats
# -------------------------
test_results <- bind_rows(shapiro_df, levene_df)

# -------------------------
# 4. Affichage table HTML avec style
# -------------------------
kable(
  test_results,
  caption = "<b>Normality and Homogeneity Tests for ANOVA</b>",
  escape = FALSE
) %>%
  table_style()

```


A Kruskal-Wallis test was conducted in lieu of ANOVA due to a violation of the normality assumption in the residuals, whereas the assumption of homogeneity of variances was satisfied.


```{r AutomationEducationTest, echo = FALSE, message = FALSE, warning = FALSE}
# ----------------------------------
# Kruskal-Wallis and Dunn Test Results with Significance
# ----------------------------------
library(dplyr)
library(knitr)
library(kableExtra)
library(FSA)        # for dunnTest

# Kruskal-Wallis test
kruskal_res <- kruskal.test(Automation_Probability_2030 ~ Education_Level, data = df)

# Post-hoc Dunn test
dunn_res <- FSA::dunnTest(Automation_Probability_2030 ~ Education_Level, data = df, method = "bonferroni")

# Extract Dunn results as a data frame and add significance
dunn_table <- dunn_res$res %>%
  rename(
    Comparison = Comparison,
    Z = Z,
    P_unadj = P.unadj,
    P_adj = P.adj
  ) %>%
  mutate(Significant = ifelse(P_adj < 0.05, "Yes", "No"))

# Display Kruskal-Wallis summary
kruskal_summary <- data.frame(
  Statistic = kruskal_res$statistic,
  df = kruskal_res$parameter,
  p_value = kruskal_res$p.value
)

kable(
  kruskal_summary,
  caption = "<b>Kruskal-Wallis Test for Automation Probability by Education Level</b>",
  digits = 4,
  escape = FALSE
) %>%
  table_style()

# Display Dunn post-hoc test with significance
kable(
  dunn_table,
  caption = "<b>Post-hoc Dunn Test (Bonferroni-adjusted) for Education Levels</b>",
  digits = 4,
  escape = FALSE
) %>%
  table_style()
```


 Both Kruskal-Wallis and Dunn tests confirm that automation probability does not significantly differ between education levels.


## Automation Probability Across Job Title


```{r anovaprobjob, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)

# -------------------------
# ANOVA Automation Probability ~ Job Title
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Job_Title, data = df)

# Résumé tidy de l'ANOVA
anova_summary <- broom::tidy(anova_model) %>%
  mutate(
    Sum_Sq = round(sumsq, 3),
    Mean_Sq = round(meansq, 3),
    F_value = round(statistic, 3),
    P_value = signif(p.value, 3)
  ) %>%
  select(term, Sum_Sq, Mean_Sq, F_value, P_value)

# Affichage avec ton style
kable(
  anova_summary,
  caption = "<b>ANOVA: Automation Probability by Job Title</b>",
  escape = FALSE
) %>%
  table_style()

```


```{r normalitejobauto, echo = FALSE, message = FALSE, warning = FALSE }
library(ggplot2)
library(gridExtra)

# -------------------------
# Modèle ANOVA
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Job_Title, data = df)
residuals_anova <- residuals(anova_model)

# -------------------------
# 1. QQ-plot des résidus pour la normalité
# -------------------------
qq_plot <- ggplot(data.frame(residuals = residuals_anova), aes(sample = residuals)) +
  stat_qq(color = "#1F4E79") +
  stat_qq_line(color = "#A61C3C", linewidth = 1) +
  labs(title = "QQ Plot of Residuals") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# -------------------------
# 2. Plot des résidus vs valeurs ajustées pour l'hétéroscédasticité
# -------------------------
fitted_values <- fitted(anova_model)
resid_plot <- ggplot(data.frame(Fitted = fitted_values, Residuals = residuals_anova), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "#1F4E79", alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))

# -------------------------
# Affichage côte à côte
# -------------------------
grid.arrange(qq_plot, resid_plot, ncol = 2)

```


```{r shapirojobauto, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(car)       # pour LeveneTest
library(knitr)
library(kableExtra)

# -------------------------
# Modèle ANOVA
# -------------------------
anova_model <- aov(Automation_Probability_2030 ~ Job_Title, data = df)
residuals_anova <- residuals(anova_model)

# -------------------------
# 1. Test de normalité des résidus (Shapiro-Wilk)
# -------------------------
shapiro_res <- shapiro.test(residuals_anova)
shapiro_df <- data.frame(
  Test = "Shapiro-Wilk (Residuals)",
  Statistic = round(shapiro_res$statistic, 3),
  P_value = signif(shapiro_res$p.value, 3)
)

# -------------------------
# 2. Test d'homogénéité des variances (Levene)
# -------------------------
# LeveneTest de car : centré sur la médiane
levene_res <- car::leveneTest(Automation_Probability_2030 ~ Job_Title, data = df, center = median)
levene_df <- data.frame(
  Test = "Levene (by Job Title)",
  Statistic = round(levene_res$`F value`[1], 3),
  P_value = signif(levene_res$`Pr(>F)`[1], 3)
)

# -------------------------
# 3. Combiner les deux résultats
# -------------------------
test_results <- bind_rows(shapiro_df, levene_df)

# -------------------------
# 4. Affichage table HTML avec style
# -------------------------
kable(
  test_results,
  caption = "<b>Normality and Homogeneity Tests for ANOVA</b>",
  escape = FALSE
) %>%
  table_style()

```


A one-way ANOVA revealed a highly significant effect of job title on automation probability but diagnostic tests indicated violations of the normality of residuals and homogeneity of variances. Consequently, a Welch ANOVA will be conducted to account for heteroscedasticity, and results will be complemented with appropriate post-hoc comparisons.


```{r welchjobautomation, echo = FALSE, message = FALSE, warning = FALSE}
# ----------------------------------
# Welch ANOVA + Games-Howell Post-hoc
# ----------------------------------

library(dplyr)
library(knitr)
library(kableExtra)
library(rstatix)

# Welch ANOVA
welch_res <- oneway.test(
  Automation_Probability_2030 ~ Job_Title,
  data = df,
  var.equal = FALSE
)

welch_table <- data.frame(
  Statistic = welch_res$statistic,
  df_num = welch_res$parameter[1],
  df_den = welch_res$parameter[2],
  p_value = welch_res$p.value
)

kable(
  welch_table,
  caption = "<b>Welch ANOVA for Automation Probability by Job Title</b>",
  digits = 4,
  escape = FALSE
) %>%
  table_style()

# Games–Howell post-hoc (top 10 most significant comparisons)
gh_res <- df %>%
  games_howell_test(
    Automation_Probability_2030 ~ Job_Title
  ) %>%
  arrange(p.adj) %>%      # sort by adjusted p-value
  slice_head(n = 10)      # keep only top 10

kable(
  gh_res,
  caption = "<b>Games–Howell Post-hoc Comparisons (Top 10)</b>",
  digits = 4,
  escape = FALSE
) %>%
  table_style()

```


A Welch ANOVA revealed a highly significant effect of job title on automation probability, F(19,1088.93) = 1609.37, p < .001, confirming substantial differences across occupations after accounting for heteroscedasticity. 

Games–Howell post-hoc comparisons indicated a clear occupational stratification of automation risk: highly skilled, knowledge-intensive professions exhibited similarly low automation probabilities, whereas routine, manual, and service-oriented occupations showed significantly higher levels of automation risk. 

Together, these findings suggest that exposure to automation is primarily structured by broad occupational task characteristics rather than by individual job titles.
 
 
## Risk Category vs. Education Level


```{r chisquare, echo = FALSE, message = FALSE, warning = FALSE}
# ==============================
# 1. Chi-Square Test : Education x Risk Category
# ==============================

library(dplyr)
library(vcd)        # pour assocstats (Cramer's V)
library(knitr)
library(kableExtra)

# Créer la table de contingence
contingency_table <- table(df$Education_Level, df$Risk_Category)

# Chi-Square test
chi_test <- chisq.test(contingency_table)

chi_summary <- data.frame(
  Statistic = chi_test$statistic,
  df = chi_test$parameter,
  p_value = chi_test$p.value,
  Cramers_V = sqrt(chi_test$statistic / (sum(contingency_table) * (min(dim(contingency_table))-1)))
)

# Affichage en style notebook
kable(
  chi_summary,
  caption = "<b>Chi-Square Test for Education Level by Risk Category + Cramer's V</b>",
  digits = 4,
  escape = FALSE
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left",
    font_size = 14
  ) %>%
  row_spec(0, bold = TRUE, background = "#1F4E79", color = "white") %>%
  column_spec(1, bold = TRUE)

```


```{r chisquaretest, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)

# Assurer facteurs
df$Education_Level <- factor(df$Education_Level)
df$Risk_Category <- factor(df$Risk_Category)

# Bar plot proportionnel
ggplot(df, aes(x = Education_Level, fill = Risk_Category)) +
  geom_bar(position = "fill", color = "black") +  # position="fill" pour proportions
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Low"="#1F4E79", "Medium"="#D55E00", "High"="#009E73")) +
  labs(
    title = "Proportion of Automation Risk Categories by Education Level",
    x = "Education Level",
    y = "Proportion",
    fill = "Risk Category"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))


```

The Kruskal-Wallis test showed no significant differences in automation probability across education levels. Consistently, the Chi-Square test on risk categories indicated no association with Education Level (χ² = 2.72, df = 6, p = 0.843; Cramer’s V = 0.021), suggesting that automation risk is largely independent of formal education.


## Predictors of Automation Probability


Given the bounded nature of the dependent variable and the strong occupational clustering observed in preliminary analyses, a mixed-effects beta regression was estimated with job title specified as a random intercept. This specification allows individual-level predictors to be assessed net of occupational-level heterogeneity.


```{r BetaRegression, echo = FALSE, message = FALSE, warning = FALSE}
# ==============================
# Mixed-Effects Beta Regression
# Automation Probability
# ==============================

library(glmmTMB)
library(dplyr)
library(knitr)
library(kableExtra)

# Ensure Job Title is a factor
df$Job_Title <- factor(df$Job_Title)

# Adjust dependent variable to stay strictly within (0,1)
epsilon <- 0.0001
df$Automation_Probability_2030_adj <-
  pmin(pmax(df$Automation_Probability_2030, epsilon), 1 - epsilon)

# Fit mixed-effects beta regression
beta_mixed_model <- glmmTMB(
  Automation_Probability_2030_adj ~
    Average_Salary +
    Years_Experience +
    AI_Exposure_Index +
    Tech_Growth_Factor +
    (1 | Job_Title),
  family = beta_family(link = "logit"),
  data = df
)

# Extract and format fixed effects
beta_mixed_summary <- summary(beta_mixed_model)$coefficients$cond %>%
  as.data.frame() %>%
  rename(
    Estimate  = Estimate,
    Std_Error = `Std. Error`,
    z_value   = `z value`,
    p_value   = `Pr(>|z|)`
  ) %>%
  mutate(
    Estimate  = round(Estimate, 4),
    Std_Error = round(Std_Error, 4),
    z_value   = round(z_value, 4),
    p_value   = round(p_value, 4)
  )

# Display results table
kable(
  beta_mixed_summary,
  caption = "<b>Mixed-Effects Beta Regression: Predictors of Automation Probability</b>",
  digits = 4,
  escape = FALSE
) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left",
    font_size = 14
  ) %>%
  row_spec(0, bold = TRUE, background = "#1F4E79", color = "white") %>%
  column_spec(1, bold = TRUE)

```


A mixed-effects beta regression was estimated with job title specified as a random intercept to account for occupational-level heterogeneity. After controlling for this clustering, none of the individual-level predictors—average salary, years of experience, AI exposure, or technological growth—were statistically significant. Coefficient estimates were small and close to zero, indicating negligible marginal effects. 

These findings suggest that variation in automation probability is largely driven by differences between occupations rather than by individual job characteristics within occupations.


```{r Caterpillarplot, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(betareg)
library(glmmTMB)
library(forcats)

# Calculer les prédictions combinées (effets fixes + aléatoires)
df$predicted_prob <- predict(beta_mixed_model, type = "response", re.form = NULL)  

# Créer un résumé par Job Title
pred_summary <- df %>%
  group_by(Job_Title) %>%
  summarise(
    mean_pred = mean(predicted_prob),
    sd_pred = sd(predicted_prob),
    .groups = 'drop'
  )

# Reordonner les métiers par probabilité prédite
pred_summary <- pred_summary %>%
  mutate(Job_Title = fct_reorder(Job_Title, mean_pred))

# Graphique
ggplot(pred_summary, aes(x = Job_Title, y = mean_pred)) +
  geom_col(fill = "#1F4E79") +
  geom_errorbar(aes(ymin = mean_pred - sd_pred, ymax = mean_pred + sd_pred),
                width = 0.3, color = "gray40") +
  coord_flip() +
  labs(
    title = "Predicted Automation Probability by Job Title",
    y = "Predicted Probability",
    x = "Job Title"
  ) +
  theme_minimal(base_size = 14)

```


This plot shows the predicted automation probability for each job title from the mixed-effects Beta regression model, with bars representing the mean predicted probability and error bars indicating variability across individuals within each job. Jobs are ordered from lowest to highest risk, highlighting which occupations are most and least exposed to automation.


## Key Takeaways: Statistical Analysis

-   **Education does not explain automation risk**: Non-parametric tests (Kruskal-Wallis, Dunn) and Chi-Square analyses show no significant differences in Automation Probability or Risk Categories by Education Level. The interaction between AI Exposure and Education is also non-significant, reinforcing the minimal role of formal education.

-   **Job title drives differences in automation risk**: Welch ANOVA and Games-Howell post-hoc tests reveal substantial variation in Automation Probability across occupations. Highly skilled, knowledge-intensive jobs show low automation risk, whereas routine/manual/service-oriented jobs show high risk.

-   **Continuous predictors show negligible effects**: Mixed-effects Beta regression with job title as a random intercept indicates that Average Salary, Years of Experience, AI Exposure, and Tech Growth have small, non-significant effects on automation probability, highlighting that individual-level predictors explain little beyond occupational clustering.

-   **Automation risk is structured by occupation, not education**: The combination of statistical tests confirms that differences in automation exposure are primarily driven by job-level characteristics rather than by individual attributes such as education or experience.